Compute Servers for Teaching Big Data

### 2025-02-19 paper.tex
.

\usepackage{setspace}
\doublespacing

@@

keywords: big data, cloud computing, data science education, programming

@@

%While teaching this course, we have to select data sets.
%Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
When teaching Stat 129, we look for data sets that satisfy the following criteria.
\begin{enumerate}
\item \textbf{Size} The files should be large enough to motivate the computational techniques we're learning, but no larger, as described above.
\item \textbf{Publicly Available} The data should be readily available for legal public download.
	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
\item \textbf{Format} 
    The data storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
    Each semester of Stat 129, we work with at least three data sets: a plain text tabular format such as CSV, a nested format such as XML or JSON, and a SQL database.
    Typically the data will be compressed, with one data set stored across multiple files.
	Avoid proprietary, novel, or highly specialized data formats.
\item \textbf{Background Knowledge}
    A college student should be able to understand what the data actually represents in a single class period.
    For example, weather data such as Global Historical Climatology Network-Daily Database \cite{menne2012overview} works great, as each row represents a weather observation for a particular date and location.
\end{enumerate}
The above criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on the specialized formats of genomic data.
%Most of these principles We've violated at one time or another, and the difficulties became clear later.
Amazon Open Data\footnote{\url{https://registry.opendata.aws/}} has provided a reliably good source of large data sets fitting these criteria. 
The US government\footnote{\url{https://data.gov/}} is another good source.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{jetstream}
    \caption{This screenshot shows how to create a new instance of a virtual machine.
The Jetstream2 user interface is far simpler and easier to teach compared to commercial cloud providers.}
    \label{fig:jetstream}
\end{figure}

@@

If it was easier to set up and run a server for teaching undergraduate data analysis classes like Stat 129, then we believe that more instructors would be interested in teaching these courses.
For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to many instructors simultaneously teaching similar courses at other institutions.
It should have the following features:
\begin{enumerate}
\item \textbf{data science tools} Students should be able to use common data science programming technogies such as R, Python, and some form of SQL.
    %One way to provide this is to give the instructor administrative privileges to install software.
    Web based tools such as Jupyter Notebooks certainly broaden the appeal and potential users of the server, but they are not strictly necessary.
\item \textbf{free or low cost} Free for students is ideal, of course.
    If free is not possible, then the cost over the semester should not exceed that of a reasonably priced textbook.
\item \textbf{instructor support}
    Instructors without a background in systems administration should be able to teach on the server, and receive some level of technical and community support.
    This is perhaps the hardest requirement, because the best way to provide dedicated support is through paid staff.
\item \textbf{long term sustainability} 
    If instructors and institutions are going to invest in creating courses that rely on a server, then they need to have some assurance that the server will continue to exist.
    A 1 year pilot program might be a useful start, but it's much more compelling to have a guarantee that a server will be available for the next 5 years.
\item \textbf{minimal specificity}
    It's ideal if students in a one semester data science class don't need to learn concepts specific to one particular system or vendor, as discussed in Sections~\ref{sec:server} and~\ref{sec:comcloud}.
\item \textbf{novice user safeguards} It should not be possible for one student to delete or damage course infrastructure, either accidentally or on purpose.
    Standard Linux user accounts without administrative privileges work fine.
\item \textbf{parallel capability}
    The parallel processing capabilities of the server should exceed those of a common laptop to provide a compelling reason to actually use the server.
    In practice, with 10 CPU's students can experience an order of magnitude speedup compared to a serial program, and this is enough to motivate the techniques in Stat 129.
    An abundance of compute power is definitely not necessary if instructors follow the advice in Section~\ref{sec:data}.
\item \textbf{shared data storage} It should be easy for an entire class to simultaneously read and share the same large data sets.
    100 GB of storage would suffice, 1 TB of storage would be great, and 10 TB would be more than enough.
\item \textbf{shell access} Students should be able to interact with their computer using bash so that they can use data processing pipelines, scripts, and version control.
\end{enumerate}

>>>>

%\usepackage{setspace}
%\doublespacing

@@

\vspace{0.5in}
keywords: big data, cloud computing, data science education, programming

@@

%While teaching this course, we have to select data sets.
%Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
When teaching Stat 129, we look for data sets that satisfy the following criteria.
\begin{enumerate}
\item \textbf{size} The files should be large enough to motivate the computational techniques we're learning, but no larger, as described above.
\item \textbf{publicly available} The data should be readily available for legal public download.
	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
\item \textbf{format} 
    The data storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
    Each semester of Stat 129, we work with at least three data sets: a plain text tabular format such as CSV, a nested format such as XML or JSON, and a SQL database.
    Typically the data will be compressed, with one data set stored across multiple files.
	Avoid proprietary, novel, or highly specialized data formats.
\item \textbf{background knowledge}
    A college student should be able to understand what the data actually represents in a single class period.
    For example, weather data such as Global Historical Climatology Network-Daily Database \cite{menne2012overview} works great, as each row represents a weather observation for a particular date and location.
\end{enumerate}
The above criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on the specialized formats of genomic data.
%Most of these principles We've violated at one time or another, and the difficulties became clear later.
Amazon Open Data\footnote{\url{https://registry.opendata.aws/}} has provided a reliably good source of large data sets fitting these criteria. 
The US government\footnote{\url{https://data.gov/}} is another good source.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{jetstream}
    \caption{This screenshot shows how to create a new instance of a virtual machine using Jetstream2.
This user interface is far simpler and easier to teach compared to commercial cloud providers.}
    \label{fig:jetstream}
\end{figure}

@@

If it was easier to set up and run a server for teaching undergraduate data analysis classes like Stat 129, then we believe that more instructors would be interested in teaching these courses.
For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to many instructors simultaneously teaching similar courses at other institutions.
It should have the following features (listed alphabetically):
\begin{enumerate}
\item \textbf{data science tools} Students should be able to use common data science programming technogies such as R, Python, and some form of SQL.
    %One way to provide this is to give the instructor administrative privileges to install software.
    Web based tools such as Jupyter Notebooks certainly broaden the appeal and potential users of the server, but they are not strictly necessary.
\item \textbf{free or low cost} Free for students is ideal, of course.
    If free is not possible, then the cost over the semester should not exceed that of a reasonably priced textbook.
\item \textbf{instructor support}
    Instructors without a background in systems administration should be able to teach on the server, and receive some level of technical and community support.
    This is perhaps the hardest requirement, because the best way to provide dedicated support is through paid staff.
\item \textbf{long term sustainability} 
    If instructors and institutions are going to invest in creating courses that rely on a server, then they need to have some assurance that the server will continue to exist.
    A 1 year pilot program might be a useful start, but it's much more compelling to have a guarantee that a server will be available for the next 5 years.
\item \textbf{minimal specificity}
    It's ideal if students in a one semester data science class don't need to learn concepts specific to one particular system or vendor, as discussed in Sections~\ref{sec:server} and~\ref{sec:comcloud}.
\item \textbf{novice user safeguards} It should not be possible for one student to delete or damage course infrastructure, either accidentally or on purpose.
    Standard Linux user accounts without administrative privileges work fine.
\item \textbf{parallel capability}
    The parallel processing capabilities of the server should exceed those of a common laptop to provide a compelling reason to actually use the server.
    In practice, with 10 CPU's students can experience an order of magnitude speedup compared to a serial program, and this is enough to motivate the techniques in Stat 129.
    An abundance of compute power is definitely not necessary if instructors follow the advice in Section~\ref{sec:data}.
\item \textbf{shared data storage} It should be easy for an entire class to simultaneously read and share the same large data sets.
    100 GB of storage would suffice, 1 TB of storage would be great, and 10 TB would be more than enough.
\item \textbf{shell access} Students should be able to interact with their computer using bash so that they can use data processing pipelines, scripts, and version control.
\end{enumerate}
